Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-06-21 2015-turklang-kazdep/paper.tex
new table
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60922 72bbbca6-d526-0410-a7d9-f06f51895060

As seen in the table, the morphologically unmarked nominative case (e.g., үй \gloss{the/a house} \gmk{nom}) has a wide variety of uses, including indefinite object and genitival.  Definite objects are marked with the accusative case (e.g., үйді \gloss{the house} \gmk{acc}), and definite genitivals are marked with the genitive case (e.g., үйдің \gloss{of the house} \gmk{gen}).  While the Kazakh transducer marks these all as \tag{nom}, their use may be disambiguated.  Genitival nominals, whether definite or not, must (NO EXCEPTIONS) have a corresponding nominal with possessive morphology that agrees in person and number (e.g., үйдің есіг\textbf{і} \gloss{the/a door to the house}), and subject nominals must have a corresponding predicate containing e.g., a verb or a copula that agrees in person and number with the nominative nominal (e.g., үй көшіріле\textbf{ді} \gloss{the house gets moved}).

@@

Another type of ``zero-derivation'' involves the use of adjectives and nouns.  In Kazakh, adjectives are all attributive, but many may also be used either nominally (i.e., as nouns) or adverbially (i.e., as adverbs), or both.  The morphological transducer internally provides a series of classes; depending on class membership, an adjective may receive the readings \tag{adj}\tag{advl} or \tag{adj}\tag{subst}, besides the default attributive reading of \tag{adj}.  If an adjective may be used substantivally, then it will receive a range of substantive morphology as well.  Similarly, many nouns may be used attributively, besides the default substantive reading.  Besides the nominative (and indefinite accusative/genitive) reading of \tag{nom}, the transducer provides an attributive reading for most nominals of \tag{nom}\tag{attr}.

>>>>

As seen in the table, the morphologically unmarked nominative case (e.g., үй \gloss{the/a house} \gmk{nom}) has a wide variety of uses, including indefinite object and genitival.  Definite objects are marked with the accusative case (e.g., үйді \gloss{the house} \gmk{acc}), and definite genitivals are marked with the genitive case (e.g., үйдің \gloss{of the house} \gmk{gen}).  The Kazakh transducer marks all bare nominals both as \tag{nom} and \tag{attr}, but the various functions may be disambiguated syntactically.  Genitival nominals, whether definite or not, must (NO EXCEPTIONS) have a corresponding nominal with possessive morphology that agrees in person and number (e.g., үйдің есіг\textbf{і} \gloss{the/a door to the house}; үй тапсырма\textbf{сы} \gloss{homework}), and subject nominals must have a corresponding predicate containing e.g., a verb or a copula that agrees in person and number with the nominative nominal (e.g., үй көшіріле\textbf{ді} \gloss{the house gets moved}).  When a nominative nominal depends on another nominal that does not have possessive case, the first one must be considered attributive (e.g., үй киім \gloss{house clothes}).

The attributive use of nominals is similar to the use of adjectives, and could even be thought of as a ``zero-derivation'' of nouns into adjectives.  Interestingly, many adjectives can also be used substantively.  Table~\ref{tab:zeroderiv} shows the various functions of nominals, adjectives, and adverbs.

\begin{table}[htbp]
	\centering
	\caption{The default (first line of each) and ``zero-derived'' uses of nominals, adjectives, and adverbs}\label{tab:zeroderiv}
	\begin{tabular}{lp{8em}p{22em}}
		\toprule
			\textbf{Category} & \textbf{Function} & \textbf{Example} \\
		\midrule
			Nominals & Substantive\newline
							Attributive\newline
							Adverbial
						& hargle\newline 
							jargal\newline
							\\
			Adjectives & Attributive\newline
							Substantive\newline
							Adverbial
						& foo \newline
						\newline
						\\
			Adverbs & Adverbial
						&
						\\
		\bottomrule
	\end{tabular}

\end{table}

Another type of ``zero-derivation'' involves the use of adjectives and nouns.  In Kazakh, adjectives are all attributive, but many may also be used either nominally (i.e., as nouns) or adverbially (i.e., as adverbs), or both.  The morphological transducer internally provides a series of classes; depending on class membership, an adjective may receive the readings \tag{adj}\tag{advl} or \tag{adj}\tag{subst}, besides the default attributive reading of \tag{adj}.  If an adjective may be used substantivally, then it will receive a range of substantive morphology as well.  Similarly, many nouns may be used attributively, besides the default substantive reading.  Besides the nominative (and indefinite accusative/genitive) reading of \tag{nom}, the transducer provides an attributive reading for most nominals of \tag{attr}.

@@
